% $Id$

\section{Implementation and Experimental Results}
\label{sec:bench}

In this section, we briefly describe the implementation of our approach
introduced previously, and present some experimental results obtained by running
this implementation over a benchmark of problems in the \bmth{} set theory.

\subsection{Implementation}

The algorithms described in this paper are implemented in the \archsat{}
automated theorem prover\footnote{Available at:
\url{https://gforge.inria.fr/projects/archsat}.}. It relies on the
\msat{}~\cite{GB17} library, derived from the \altergoz{} tool, and which is a
generic library for building automated deduction tools based on SAT solvers. The
SAT core in \msat{} is implemented using CDCL rather than the DPLL strategy
described in Sec.~\ref{sec:smt}, but our approach is actually agnostic with
respect to the SAT solver core and it therefore does not have any impact on the
first order reasoning or rewriting presented previously. \archsat{} (as well as
\msat{}) is written in \ocaml{}. \archsat{} natively supports polymorphic types
as described in~\cite{BP13}.

\subsection{Experimental Results}

As a framework to test our tool, we consider the set theory of the \bmth{}
method~\cite{B-Book}. This method is supported by some tool sets, such as
\atelierb{}, which are used in industry to specify and build, by stepwise
refinements, software that is correct by design. This theory is suitable as it
can be easily turned into a theory that is compatible with deduction modulo
theory, i.e. where a large part of axioms can be turned into rewrite rules, and
for which the rewriting theory proposed previously in Subsec.~\ref{sec:rew}
should work. Starting from the theory described in Chap.~2 of the
\bbook{}~\cite{B-Book}, we therefore transform whenever possible the axioms and
definitions into rewrite rules. The resulting theory has been introduced
in~\cite{BA15}, and due to lack of space, we only provide, in
Fig.~\ref{fig:bset}, the three rewriting rules corresponding to the axiomatic
core of the \bmth{} set theory that we consider.

As can be seen, the proposed theory is typed, using first order logic extended
to polymorphic types à la ML, through a type system in the spirit
of~\cite{BP13}. This extension to polymorphic types offers more flexibility, and
in particular allows us to deal with theories that rely on elaborate type
systems, like the \bmth{} set theory (see Chap.~2 of the
\bbook{}~\cite{B-Book}). The complete type system that is used in this
formalization can be found in~\cite{BA15}. The type constructors,
i.e. $\mathsf{tup}$ for tuples and $\mathsf{set}$ for sets, and type schemes of
the considered set constructs are provided in Fig.~\ref{fig:bset} as well, where
\type{} is the type of types and $\omicron$ the type of formulas, and where type
arguments are subscript annotations of the constructs.

\begin{figure}[t]
\small
\hspace{0.2cm}\underline{Axioms of Set Theory}
\begin{flushleft}
$\begin{array}{@{\hspace{0.2cm}}l}
(x,y)_{\alpha_1,\alpha_2}\in_{\tuple{\alpha_1}{\alpha_2}}s\times_{\alpha_1,\alpha_2}t\rew
x\in_{\alpha_1}s\land{}y\in_{\alpha_2}t\\
s\in_{\set{\alpha}}\mathbb{P}_\alpha(t)\rew
\forall{}x:\alpha.x\in_\alpha{}s\Rightarrow{}x\in_\alpha{}t\\
s=_{\set{\alpha}}t\rew
\forall{}x:\alpha.x\in_\alpha{}s\Leftrightarrow{}x\in_\alpha{}t
\end{array}$
\end{flushleft}
\hspace{0.2cm}\underline{Type Constructors}
\begin{flushleft}
$\begin{array}{@{\hspace{0.2cm}}l@{\hspace{1.0cm}}l}
\mathsf{tup}:\Pi\alpha_1,\alpha_2:\type.\type &
\mathsf{set}:\Pi\alpha:\type.\type
\end{array}$
\end{flushleft}
\hspace{0.2cm}\underline{Type Schemes of the Set Constructs}
\begin{flushleft}
$\begin{array}{@{\hspace{0.2cm}}lcl}
\arg\in\arg & : &
\Pi\alpha:\type.\alpha\rightarrow\set{\alpha}\rightarrow\omicron\\
(\arg,\arg) & : &
\Pi\alpha_1,\alpha_2:\type.\alpha_1\rightarrow\alpha_2\rightarrow
\tuple{\alpha_1}{\alpha_2}\\
\arg\times\arg & : &
\Pi\alpha_1,\alpha_2:\type.\set{\alpha_1}\rightarrow\set{\alpha_2}\rightarrow
\set{\tuple{\alpha_1}{\alpha_2}}\\
\mathbb{P}(\arg) & : &
\Pi\alpha:\type.\set{\alpha}\rightarrow\set{\set{\alpha}}\\
\arg=\arg & : & \Pi\alpha:\type.\alpha\rightarrow\alpha\rightarrow\omicron\\
\end{array}$
\end{flushleft}
\caption{Rewriting Rules of the Axiomatic Core of the \bmth{} Set Theory}
\label{fig:bset}
\end{figure}

To test \archsat{} in this theory, we consider 319~lemmas coming from Chap.~2 of
the \bbook{}~\cite{B-Book}. These lemmas are properties of various difficulty
regarding the set constructs introduced by the \bmth{} method. It should be
noted that these constructs and notations are, for a large part of them,
specific to the \bmth{} method, as they are used for the modeling of industrial
projects, and are not necessarily standard in set theory.

As tools, we consider \archsat{} (development version\footnote{\git{}
version~7720d8c.}). We also include other automated theorem provers, able to
deal with first order logic with polymorphic types and rewriting natively. In
particular, we consider \zipper{} (version~1.5), a prover based on superposition
and rewriting, as well as \zenm{} (version~0.4.2), a tableau-based prover that
is an extension of \zenon{} to deduction modulo theory. To show the impact of
rewriting over the results, we also include the \altergo{} SMT solver
(version~1.01). It would have been possible to also consider provers dealing
with pure first order logic and encode the polymorphic layer. But preliminary
tests have been conducted and very low results have been obtained even for the
best state-of-the-art provers (we have considered \e{} and \cvc{} in
particular), which tends to show that polymorphism encoding adds a lot of noise
in proof search and is not effective in practice.

The experiment was run on an \intel{}~3.50~GHz computer, with a timeout of 90~s
(beyond this timeout, results do not change) and a memory limit of 1~GiB. The
results are summarized in Tab.~\ref{tab:bench}. In these results, we observe
that \archsat{} obtains better results, in terms of proved problems, than
\zenm{} and \altergo{}, which tends to show the effectiveness of our approach in
practice. However, \archsat{} places second behind \zipper{}, which means that
there is still room for improvement regarding the implementation of our tool.
Looking at the cumulative times, \altergo{} is not really faster than
\archsat{} and \zipper{}, which take more time to find few more difficult
problems (with a timeout of 3~s, they respectively find 260 and 303~proofs in
16.61~s and 17.61~s, while \altergo{} obtains the same results).

\setlength{\tabcolsep}{3pt}
\renewcommand{\arraystretch}{1.2}
\newcolumntype{C}{>{\centering}X}

\begin{table}[t]
\begin{center}
\begin{tabularx}{\textwidth}{|X|C|C|C|C|}
\hline
\begin{tabular}{l}
319~Problems
\end{tabular} &
\begin{tabular}{c}
\archsat{}
\end{tabular} &
\begin{tabular}{c}
\zipper{}
\end{tabular} &
\begin{tabular}{c}
\zenm{}
\end{tabular} &
\begin{tabular}{c}
\altergo
\end{tabular}\tabularnewline
\hline
\begin{tabular}{l}
Proofs
\end{tabular} &
\begin{tabular}{c}
272
\end{tabular} &
\begin{tabular}{c}
306
\end{tabular} &
\begin{tabular}{c}
138
\end{tabular} &
\begin{tabular}{c}
232
\end{tabular}\tabularnewline
\hline
\begin{tabular}{l}
Rate
\end{tabular} &
\begin{tabular}{c}
85.3\%
\end{tabular} &
\begin{tabular}{c}
95,9\%
\end{tabular} &
\begin{tabular}{c}
43.3\%
\end{tabular} &
\begin{tabular}{c}
72.7\%
\end{tabular}\tabularnewline
\hline
\begin{tabular}{l}
Time \small{(s)}
\end{tabular} &
\begin{tabular}{c}
268.69
\end{tabular} &
\begin{tabular}{c}
109.88
\end{tabular} &
\begin{tabular}{c}
2.86
\end{tabular} &
\begin{tabular}{c}
8.42
\end{tabular}\tabularnewline
\hline
\end{tabularx}
\end{center}
\caption{Experimental Results over the \bmth{} Set Theory Benchmark}
\label{tab:bench}
\end{table}

\renewcommand{\arraystretch}{1}
