% $Id$

\section{SAT Solving Modulo Tableau Theory}
\label{sec:smt}

\todo{introduction, also explain why this is better than a
  regular (propositional) Tableau prover, and what the advantages are}
\todo{cite Satallax, brief comparison (w.r.t quantifiers)}

\subsection{SAT Solving Modulo Theories}

In this section, we recall the classic architecture of SMT
solvers~\cite{barrett2006splitting}.  We introduce $\mathcal{T}$ and
$\mathcal{F}$ respectively the sets of first order terms and first order
formulas over the signature
$\mathcal{S}=(\mathcal{S}_\mathcal{F},\mathcal{S}_\mathcal{P})$, where
$\mathcal{S}_\mathcal{F}$ is the set of function symbols, and
$\mathcal{S}_\mathcal{P}$, the set of predicate symbols, such that
$\mathcal{S}_\mathcal{F}\cap\mathcal{S}_\mathcal{P}=\emptyset$. The set
$\mathcal{T}$ is extended with two kinds of terms specific to tableau proof
search, i.e. $\epsilon{}$-terms (used instead of Skolemization) of the form
$\epsilon(x).P(x)$, where $P(x)$ is a formula, and which means some $x$ that
satisfies $P(x)$, if it exists, and metavariables (often named free variables in
the tableau-related literature) of the form $X_P$, where $P$ is the formula that
introduces the metavariable, and which is either $\forall{}x.Q(x)$ or
$\neg\exists{}x.Q(x)$, with $Q(x)$ a formula.

A boxed formula is of the form $\lfloor{}P\rfloor$, where $P$ is a formula. A
boxed formula is called atom, and a literal is either an atom, or the negation
of an atom. A literal is such that there is no negation on top of the boxed
formula (which means that $\lfloor\neg{}P\rfloor=\neg\lfloor{}P\rfloor$, and
that $\neg\neg\lfloor{}P\rfloor=\lfloor{}P\rfloor$). A clause is a disjunction
of literals. It should be noted that SAT solving usually reasons over sets of
clauses composed of first order literals; here, a literal is a first order
formula (possibly with quantifiers), which requires to box formulas to get a
regular SAT solving problem where boxed formulas are propositional variables.

An assignment of the boxed formulas is a partial function that assigns to boxed
formulas an element of $\{\top,\bot\}$ (values of the Boolean algebra). Given a
clause $C$, we say that $C$ is valid and write $\models{}C$, if $C$ evaluates to
$\top$ for every assignment. Given a set of clauses $S$, we say that $S$ is
valid and write $\models{}S$, if each clause of $S$ is valid. Given two sets of
clauses $S$ and $S'$, we say that $S$ implies $S'$ and write $S\models{}S'$ if,
for every assignment, if each clause of $S$ evaluates to $\top$ then each clause
of $S'$ evaluates to $\top$. If $S$ is a set of literals, then we say that $S$
is a model of $S'$.

A theory $T$ is a set of formulas over a given signature $\mathcal{S}$, which
are called axioms of the theory $T$. A clause $C$ is a tautology in the theory
$T$, written $\models_TC$, if the axioms of $T$ imply the formula $C'$, where
$C'$ is the clause $C$ where all the boxed formulas have been unboxed. A set of
clauses $S$ is a tautology in the theory $T$ if each clause of $S$ is a
tautology in $T$. Given two sets of clauses $M$ and $S$, we say that $M$ is a
model of $S$ modulo the theory $T$ and write $M\models_TS$, if $M$ is a model of
$S$ and $M$ is a tautology in $T$.

A SAT solver modulo theories is characterized by a set of theories $S_T$ and an
internal state of the form $M\parallel{}S$, where $M$ is an ordered list of
literals (if $M=l_1,l2,\ldots,l_n$ then $l_1<l_2<\ldots<l_n$) and $S$ a set of
clauses. Intuitively, $M$ represents the model of $S$, which will be built
progressively (initially $M=\emptyset$) by the application of a set of rules
over the internal state of the solver. This set of rules is presented in
Fig.~\ref{fig:smt}. Each time a literal is propagated (rule
``$\mathrm{unit~prop}$'') or decided (rule ``$\mathrm{decide}$''), it is given
to each theory, which may generate new clauses (rule ``$\mathrm{learn}$''). If a
clause is unvalidated by the model, then it is possible to backtrack the
decision over a given literal of the model if it exists (rule
``$\mathrm{backjump}$'').  There are two final states in this transition
system. The first one is ``$\mathrm{unsat}$'', which means that $S$ is
unsatisfiable (rule ``$\mathrm{unsat}$''). The second one is $M\parallel{}S$,
where $M$ is a model of $S$ and a tautology for every theory in $S_T$. In this
last case, if $M$ is not a tautology in a theory $T$, this theory generates a
set of conflict clauses (rule ``$\mathrm{learn}$'').

\begin{figure}[t]
\parbox{\textwidth}
{\small
\begin{center}
$\begin{array}{l@{\hspace{0.4cm}}l@{\hspace{0.4cm}}l}
M\parallel{}S,C\lor{}l\longrightarrow{}M,l\parallel{}S,C\lor{}l &
(\mathrm{unit~prop}) & \mbox{if }l\not\in{}M\mbox{ and }M\models{}\neg{}C\\\\

M\parallel{}S\longrightarrow{}M,l^\mathrm{d}\parallel{}S & (\mathrm{decide}) &
\mbox{if }l\not\in{}M\mbox{, and }l\in{}S\mbox{ or }\neg{}l\in{}S\\\\

M\parallel{}S,C\longrightarrow\mathrm{unsat} & (\mathrm{unsat}) &
\mbox{if }M'\models{}\neg{}C\mbox{ s.t. }M'\subseteq{}M\\
&& \mbox{and there is no }l^\mathrm{d}\leq{}l'\mbox{ in }M\\
&& \mbox{for all }l'\in{}M'\\\\
 
M,l^\mathrm{d},M'\parallel{}S,C\longrightarrow{}M,l'\parallel{}S,C &
(\mathrm{backjump}) & \mbox{if }M,l^\mathrm{d},M'\models\neg{}C
\mbox{, and there is}\\
&& \mbox{some clause }C'\lor{}l' s.t.:\\
&& l'\not\in{}M\mbox{, and }l'\in{}S\mbox{ or }\neg{}l'\in{}S\\
&& \mbox{or }l'\in{}M,l^\mathrm{d},M'\mbox{ or }
\neg{}l'\in{}M,l^\mathrm{d},M',\\
&& \mbox{and }S,C\models{}C'\lor{}l'\mbox{, and }M\models\neg{}C'\\\\

M\parallel{}S\longrightarrow{}M\parallel{}S,S'
& (\mathrm{learn}) & \models_TS'\mbox{, where }T\mbox{ is a theory}
\end{array}$
\end{center}}
\caption{Rules of SAT Solving Modulo Theory}
\label{fig:smt}
\end{figure}

\subsection{The Tableau Theory}

In the SAT solver previously described, the tableau proof search method is
integrated as a regular theory. When a literal is propagated or decided, we
generate a set of clauses corresponding to the application of a tableau rule
over the logical connective at root of the formula in the box of the
literal. More precisely, for a literal $l$, we generate the set of clauses
$\llbracket{}l\rrbracket$, where the function $\llbracket\cdot\rrbracket$ is
described by the rules of Fig.~\ref{fig:tabth}. When a literal is propagated or
decided, we use all the rules of Fig.~\ref{fig:tabth} except the instantiation
$\gamma$-rules (rules $\gamma_{\forall\mathrm{inst}}$ and
$\gamma_{\neg\exists\mathrm{inst}}$). It should be noted that we use the same
names for the rules than in tableau calculus ($\alpha$-rules, $\beta$-rules,
etc.), but there is no precedence between rules and therefore no priority in the
application of the rules contrary to the tableau proof search method (where
$\alpha$ rules are applied before $\beta$-rules, and so on).

When the SAT solver reaches a state $M\parallel{}S$, where $M$ is a model of
$S$, we look for a conflict in $M$ between two literals by unification and we
generate the clauses corresponding to the instantiation of the metavariables
using the result of the unification. More precisely, if there exist two literals
$l$ and $\neg{}l'$ in $M$ such that $l=\lfloor{}Q\rfloor$ and
$l'=\lfloor{}R\rfloor$, with $Q$ and $R$ two formulas, then for each
substitution $(X_{\forall{}x.P(x)}\mapsto{}t)\in\mathrm{mgu}(Q,R)$ (resp.
$(X_{\neg\exists{}x.P(x)}\mapsto{}t)\in\mathrm{mgu}(Q,R)$) such that there is no
$Y_T\in\forall{}x.P(x)$ (resp. $Y_T\in\neg\exists{}x.P(x)$), we can generate the
clauses $\llbracket\lfloor\forall{}x.P(x)\rfloor{}\rrbracket$ 
(resp. $\llbracket\neg\lfloor\exists{}x.P(x)\rfloor{}\rrbracket$) using the rule
$\gamma_{\forall\mathrm{inst}}$ (resp. $\gamma_{\neg\exists\mathrm{inst}}$) of
Fig.~\ref{fig:tabth}.

\begin{figure}[t]
\parbox{\textwidth}
{\small
\underline{Analytic Rules}
\begin{center}
$\begin{array}{lll@{\hspace{0.5cm}}l}
\llbracket\lfloor{}P\land{}Q\rfloor{}\rrbracket & = &
\neg\lfloor{}P\land{}Q\rfloor\lor\lfloor{}P\rfloor,
\neg\lfloor{}P\land{}Q\rfloor\lor\lfloor{}Q\rfloor & (\alpha_\land)\\\\

\llbracket\neg\lfloor{}P\land{}Q\rfloor{}\rrbracket & = &
\lfloor{}P\land{}Q\rfloor\lor\neg\lfloor{}P\rfloor\lor\neg\lfloor{}Q\rfloor &
(\beta_{\neg\land})\\\\

\llbracket{}\lfloor{}P\lor{}Q\rfloor{}\rrbracket & = &
\neg\lfloor{}P\lor{}Q\rfloor\lor\lfloor{}P\rfloor\lor\lfloor{}Q\rfloor &
(\beta_\lor)\\\\

\llbracket\neg\lfloor{}P\lor{}Q\rfloor{}\rrbracket & = &
\lfloor{}P\lor{}Q\rfloor\lor\neg\lfloor{}P\rfloor,
\lfloor{}P\lor{}Q\rfloor\lor\neg\lfloor{}Q\rfloor &
(\alpha_{\neg\lor})\\\\

\llbracket{}\lfloor{}P\Rightarrow{}Q\rfloor{}\rrbracket & = &
\neg\lfloor{}P\Rightarrow{}Q\rfloor\lor\neg\lfloor{}P\rfloor\lor
\lfloor{}Q\rfloor & (\beta_\Rightarrow)\\\\

\llbracket\neg\lfloor{}P\Rightarrow{}Q\rfloor{}\rrbracket & = &
\lfloor{}P\Rightarrow{}Q\rfloor\lor\lfloor{}P\rfloor,
\lfloor{}P\Rightarrow{}Q\rfloor\lor\neg\lfloor{}Q\rfloor &
(\alpha_{\neg\Rightarrow})\\\\

\llbracket{}\lfloor{}P\Leftrightarrow{}Q\rfloor{}\rrbracket & = &
\neg\lfloor{}P\Leftrightarrow{}Q\rfloor\lor\lfloor{}P\Rightarrow{}Q\rfloor,
\neg\lfloor{}P\Leftrightarrow{}Q\rfloor\lor\lfloor{}Q\Rightarrow{}P\rfloor &
(\beta_\Rightarrow)\\\\

\llbracket{}\neg\lfloor{}P\Leftrightarrow{}Q\rfloor{}\rrbracket &
= &
\lfloor{}P\Leftrightarrow{}Q\rfloor\lor\neg\lfloor{}P\Rightarrow{}Q\rfloor\lor
\neg\lfloor{}Q\Rightarrow{}P\rfloor & (\beta_{\neg\Rightarrow})
\end{array}$
\end{center}

\underline{$\delta$-Rules}
\begin{center}
$\begin{array}{lll@{\hspace{0.5cm}}l}
\llbracket\lfloor\exists{}x.P(x)\rfloor{}\rrbracket & = &
\neg{}\lfloor\exists{}x.P(x)\rfloor\lor\lfloor{}P(\epsilon(x).P(x))\rfloor &
(\delta_\exists)\\\\

\llbracket\neg\lfloor\forall{}x.P(x)\rfloor{}\rrbracket & = &
\lfloor\forall{}x.P(x)\rfloor\lor\neg\lfloor{}P(\epsilon(x).\neg{}P(x))\rfloor &
(\delta_{\neg\forall})
\end{array}$
\end{center}

\underline{$\gamma$-Rules}
\begin{center}
$\begin{array}{lll@{\hspace{0.5cm}}l}
\llbracket\lfloor\forall{}x.P(x)\rfloor{}\rrbracket & = &
\neg\lfloor\forall{}x.P(x)\rfloor\lor\lfloor{}P(X_{\forall{}x.P(x)})\rfloor &
(\gamma_{\forall{}M})\\\\

\llbracket\neg\lfloor\exists{}x.P(x)\rfloor{}\rrbracket & = &
\lfloor\exists{}x.P(x)\rfloor\lor
\neg\lfloor{}P(X_{\neg\exists{}x.P(x)})\rfloor &
(\gamma_{\neg\exists{}M})\\\\

\llbracket\lfloor\forall{}x.P(x)\rfloor{}\rrbracket & = &
\neg\lfloor\forall{}x.P(x)\rfloor\lor\lfloor{}P(t)\rfloor &
(\gamma_{\forall\mathrm{inst}})\\\\

\llbracket\neg\lfloor\exists{}x.P(x)\rfloor{}\rrbracket & = &
\lfloor\exists{}x.P(x)\rfloor\lor\neg\lfloor{}P(t)\rfloor &
(\gamma_{\neg\exists\mathrm{inst}})
\end{array}$
\end{center}}
\caption{Rules of Tableau Theory}
\label{fig:tabth}
\end{figure}

To show how this theory works, let us prove that
$\exists{}x.P(x)\Rightarrow{}P(a)\land{}P(b)$, where $P$ is a predicate symbol
and $a,b$ two constants. The SAT solver is initiated with the state
$\emptyset\parallel
\neg\lfloor\exists{}x.P(x)\Rightarrow{}P(a)\land{}P(b)\rfloor$, and the proof is
described in Fig.~\ref{fig:tab}, where $X$ is a shortcut for
$X_{\neg\exists{}x.P(x)\Rightarrow{}P(a)\land{}P(b)}$. It should be noted that
to do this proof in sequent calculus, a right contraction is necessary to
instantiate the formula twice (with $a$ and $b$). In Fig.~\ref{fig:tab}, it is
done by propagating the literal $\lfloor{}P(X)\rfloor{}$, which can provide as
many instantiations (by unification) as necessary.

\begin{figure}[t!]
\parbox{\textwidth}
{\small
\begin{center}
$\begin{array}{lcl}
\emptyset\parallel
\neg(A\equiv\lfloor\exists{}x.P(x)\Rightarrow{}P(a)\land{}P(b)\rfloor) &
\longrightarrow & (\mathrm{unit~prop})\\

\neg{}A\parallel{}\neg{}A & \longrightarrow & (\mathrm{learn})\\

\neg{}A\parallel
\neg{}A,A\lor\neg(B\equiv\lfloor{}P(X)\Rightarrow{}P(a)\land{}P(b)\rfloor) &
\longrightarrow & (\mathrm{unit~prop})\\

\neg{}A,\neg{}B\parallel\neg{}A,A\lor\neg{}B & \longrightarrow &
(\mathrm{learn})\\

\neg{}A,\neg{}B\parallel\neg{}A,A\lor\neg{}B,
B\lor{}(C\equiv\lfloor{}P(X)\rfloor{}),\\
~~~~B\lor\neg(D\equiv\lfloor{}P(a)\land{}P(b)\rfloor{}) &
\longrightarrow & (\mathrm{unit~prop})\times{}2\\

\neg{}A,\neg{}B,C,\neg{}D\parallel\neg{}A,A\lor\neg{}B,B\lor{}C,B\lor\neg{}D &
\longrightarrow & (\mathrm{learn})\\

\neg{}A,\neg{}B,C,\neg{}D\parallel\neg{}A,A\lor\neg{}B,B\lor{}C,B\lor\neg{}D,\\
~~~~D\lor\neg(E\equiv\lfloor{}P(a)\rfloor)\lor
\neg(F\equiv\lfloor{}P(b)\rfloor) & \longrightarrow & (\mathrm{decide})\\

\neg{}A,\neg{}B,C,\neg{}D,\neg{}E^d\parallel
\neg{}A,A\lor\neg{}B,B\lor{}C, & \longrightarrow &
(\mathrm{learn})\\
~~~~B\lor\neg{}D,D\lor\neg{}E\lor\neg{}F && \{X\mapsto{}a\}=\mathrm{mgu}(C,E)\\

\neg{}A,\neg{}B,C,\neg{}D,\neg{}E^d\parallel
\neg{}A,A\lor\neg{}B,B\lor{}C,\\
~~~~B\lor\neg{}D,D\lor\neg{}E\lor\neg{}F,\\
~~~~A\lor\neg(G\equiv{}\lfloor{}P(a)\Rightarrow{}P(a)\land{}P(b)\rfloor) &
\longrightarrow & (\mathrm{unit~prop})\\

\neg{}A,\neg{}B,C,\neg{}D,\neg{}E^d,\neg{}G\parallel
\neg{}A,A\lor\neg{}B,B\lor{}C,\\
~~~~B\lor\neg{}D,D\lor\neg{}E\lor\neg{}F,A\lor\neg{}G & \longrightarrow &
(\mathrm{learn})\\

\neg{}A,\neg{}B,C,\neg{}D,\neg{}E^d,\neg{}G\parallel
\neg{}A,A\lor\neg{}B,B\lor{}C,\\
~~~~B\lor\neg{}D,D\lor\neg{}E\lor\neg{}F,A\lor\neg{}G,G\lor{}E,G\lor\neg{}D &
\longrightarrow & (\mathrm{backjump})\\

\neg{}A,\neg{}B,C,\neg{}D,E\parallel
\neg{}A,A\lor\neg{}B,B\lor{}C,\\
~~~~B\lor\neg{}D,D\lor\neg{}E\lor\neg{}F,A\lor\neg{}G,G\lor{}E,G\lor\neg{}D &
\longrightarrow & (\mathrm{unit~prop})\times{}2\\

\neg{}A,\neg{}B,C,\neg{}D,E,\neg{}F,\neg{}G\parallel
\neg{}A,A\lor\neg{}B,B\lor{}C, & \longrightarrow & (\mathrm{learn})\\
~~~~B\lor\neg{}D,D\lor\neg{}E\lor\neg{}F,A\lor\neg{}G,G\lor{}E,G\lor\neg{}D &&
\{X\mapsto{}b\}=\mathrm{mgu}(C,F)\\

\neg{}A,\neg{}B,C,\neg{}D,E,\neg{}F,\neg{}G\parallel
\neg{}A,A\lor\neg{}B,B\lor{}C,\\
~~~~B\lor\neg{}D,D\lor\neg{}E\lor\neg{}F,A\lor\neg{}G,G\lor{}E,G\lor\neg{}D\\
~~~~A\lor\neg(H\equiv{}\lfloor{}P(b)\Rightarrow{}P(a)\land{}P(b)\rfloor) &
\longrightarrow & (\mathrm{unit~prop})\\

\neg{}A,\neg{}B,C,\neg{}D,E,\neg{}F,\neg{}G,\neg{}H\parallel
\neg{}A,A\lor\neg{}B,B\lor{}C,\\
~~~~B\lor\neg{}D,D\lor\neg{}E\lor\neg{}F,A\lor\neg{}G,G\lor{}E,G\lor\neg{}D\\
~~~~A\lor\neg{}H & \longrightarrow & (\mathrm{learn})\\

\neg{}A,\neg{}B,C,\neg{}D,E,\neg{}F,\neg{}G,\neg{}H\parallel
\neg{}A,A\lor\neg{}B,B\lor{}C,\\
~~~~B\lor\neg{}D,D\lor\neg{}E\lor\neg{}F,A\lor\neg{}G,G\lor{}E,G\lor\neg{}D\\
~~~~A\lor\neg{}H,H\lor{}F,H\lor{}\neg{}D & \longrightarrow &
(\mathrm{unsat})\\

\mathrm{unsat}
\end{array}$
\end{center}}
\caption{Example of Proof Using the Tableau Theory}
\label{fig:tab}
\end{figure}

\subsection{The Rewriting Theory}

A rewriting theory allows us to introduce a computational behavior to the SAT
solver. We aim to integrate rewriting in the broadest sense of the term as
proposed by deduction modulo theory. Deduction modulo theory~\cite{DA03} focuses
on the computational part of a theory, where axioms are transformed into rewrite
rules, which induces a congruence over propositions, and where reasoning is
performed modulo this congruence. In deduction modulo theory, this congruence is
then induced by a set of rewrite rules over both propositions and terms.

In the following, we borrow some of the notations and definitions
of~\cite{DA03}. We call $\mathrm{FV}$ the function that returns the set of
free variables of a term or a formula. A term rewrite rule is a pair of terms
denoted by $l\rew{}r$, where $\mathrm{FV}(r)\subseteq\mathrm{FV}(l)$. A
proposition rewrite rule is a pair of propositions denoted by $l\rew{}r$, where
$l$ is an atomic proposition and $r$ is an arbitrary proposition, and where
$\mathrm{FV}(r)\subseteq\mathrm{FV}(l)$. A class rewrite system is a pair of
rewrite systems, denoted by $\mathcal{RE}$, consisting of $\mathcal{R}$, a set
of proposition rewrite rules, and $\mathcal{E}$, a set of term rewrite rules.

Given a class rewrite system $\mathcal{RE}$, the relations $=_\mathcal{E}$ and
$=_\mathcal{RE}$ are the congruences generated respectively by the sets
$\mathcal{E}$ and $\mathcal{R}\cup\mathcal{E}$. In the following, we use the
standard concepts of subterm and term replacement: given an occurrence $\omega$
in a proposition $P$, we write $P_{|\omega}$ for the term or proposition at
$\omega$, and $P[t]_\omega$ for the proposition obtained by replacing
$P_{|\omega}$ by $t$ in $P$ at $\omega$. Given a class rewrite system
$\mathcal{RE}$, the proposition $P$ $\mathcal{RE}$-rewrites to $P'$, denoted by
$P\rew_\mathcal{RE}P'$, if $P=_\mathcal{E}Q$, $Q_{|\omega}=\sigma(l)$, and
$P'=_\mathcal{E}Q[\sigma(r)]_\omega$, for some rule $l\rew{}r\in\mathcal{R}$,
some proposition $Q$, some occurrence $\omega$ in $Q$, and some substitution
$\sigma$.

The relation $=_\mathcal{RE}$ is not decidable in general, but there are some
cases where this relation is decidable depending on the class rewrite system
$\mathcal{RE}$ and the rewrite relation $\rew_\mathcal{RE}$. In particular, if
the rewrite relation $\rew_\mathcal{RE}$ is confluent and (weakly) terminating,
then the relation $=_\mathcal{RE}$ is decidable.

The rewriting theory is integrated into the SAT solver in a similar way than for
the tableau theory. When a literal is propagated or decided, we generate a
clause introducing a new formula that is congruent to the formula in the box of
the literal. More precisely, given a literal $\lfloor{}P\rfloor$, where $P$ is a
formula, and a formula $P'$ such that $P=_\mathcal{RE}P'$, we generate the
following clause:

$$(\bigvee_{(l,r)\in\mathcal{R}}
\neg\lfloor\forall{}\vec{x}.l\Leftrightarrow{}r\rfloor)\lor
(\bigvee_{(l,r)\in\mathcal{E}}\neg\lfloor\forall{}\vec{x}.l=r\rfloor)\lor
\lfloor{}P\Leftrightarrow{}P'\rfloor$$

where $\vec{x}=\mathrm{FV}(l)\cup\mathrm{FV}(r)$.

It should be noted that in usual SMT solvers, rewriting can be emulated by means
of triggers that are actually the left-hand side members of the class rewrite
system $\mathcal{RE}$ introduced above. But, in our rewriting theory, we can
generate the formula where rewritings have been done, while triggers can just
generate bindings, i.e. instances of the rewrite rules, which are used later to
relate the initial formula and the rewritten formula. Moreover, in our case, we
can perform several rewritings at once, while a trigger can only emulate one
rewriting.

Let us illustrate the use of the rewriting theory by means of an example in set
theory. Let us prove that
$(\forall{}s,t.s\subseteq{}t\Leftrightarrow{}\forall{}x.x\in{}s\Rightarrow{}
x\in{}t)\Rightarrow{}a\subseteq{}a$, where $a$ is a constant. The proof is given
by Fig.~\ref{fig:rew}, where the rules ``$\mathrm{learn~tab}$'' and
``$\mathrm{learn~rew}$'' represent the rule ``$\mathrm{learn}$'' using
respectively the tableau and rewriting theories.

\begin{figure}[t!]
\parbox{\textwidth}
{\small
\begin{center}
$\begin{array}{lcl}
\emptyset\parallel
\neg(A\equiv\lfloor(\forall{}s,t.s\subseteq{}t\Leftrightarrow{}
\forall{}x.x\in{}s\Rightarrow{}x\in{}t)\Rightarrow{}a\subseteq{}a\rfloor) &
\longrightarrow & (\mathrm{unit~prop})\\

\neg{}A\parallel{}\neg{}A & \longrightarrow & (\mathrm{learn~tab})\\

\neg{}A\parallel
\neg{}A,A\lor(B\equiv\lfloor\forall{}s,t.s\subseteq{}t\Leftrightarrow{}
\forall{}x.x\in{}s\Rightarrow{}x\in{}t\rfloor),\\
~~~~A\lor\neg(C\equiv\lfloor{}a\subseteq{}a\rfloor) &
\longrightarrow & (\mathrm{unit~prop})\times{}2\\

\neg{}A,B,\neg{}C\parallel\neg{}A,A\lor{}B,A\lor\neg{}C &
\longrightarrow & (\mathrm{learn~rew})\\

\neg{}A,B,\neg{}C\parallel\neg{}A,A\lor{}B,A\lor\neg{}C,\\
~~~~\neg{}B\lor(D\equiv\lfloor{}a\subseteq{}a\Leftrightarrow
\forall{}x.x\in{}a\Rightarrow{}x\in{}a\rfloor) &
\longrightarrow & (\mathrm{unit~prop})\\

\neg{}A,B,\neg{}C,D\parallel\neg{}A,A\lor{}B,A\lor\neg{}C,\neg{}B\lor{}D &
\longrightarrow & (\mathrm{learn~tab})\\

\neg{}A,B,\neg{}C,D\parallel\neg{}A,A\lor{}B,A\lor\neg{}C,\neg{}B\lor{}D,\\
~~~~\neg{}D\lor(E\equiv\lfloor{}a\subseteq{}a\Rightarrow
\forall{}.x\in{}a\Rightarrow{}x\in{}a\rfloor),\\
~~~~\neg{}D\lor(F\equiv\lfloor(\forall{}.x\in{}a\Rightarrow{}x\in{}a)
\Rightarrow{}a\subseteq{}a\rfloor) & \longrightarrow &
(\mathrm{unit~prop})\times{}2\\

\neg{}A,B,\neg{}C,D,E,F\parallel\neg{}A,A\lor{}B,A\lor\neg{}C,\neg{}B\lor{}D,
\neg{}D\lor{}E,\\
~~~~\neg{}D\lor{}F & \longrightarrow & (\mathrm{learn~tab})\\

\neg{}A,B,\neg{}C,D,E,F\parallel\neg{}A,A\lor{}B,A\lor\neg{}C,\neg{}B\lor{}D,
\neg{}D\lor{}E\\
~~~~\neg{}D\lor{}F,
\neg{}F\lor\neg(G\equiv\lfloor\forall{}.x\in{}a\Rightarrow{}x\in{}a\rfloor)
\lor{}C & \longrightarrow & (\mathrm{unit~prop})\\

\neg{}A,B,\neg{}C,D,E,F,\neg{}G\parallel\neg{}A,A\lor{}B,A\lor\neg{}C,
\neg{}B\lor{}D,\neg{}D\lor{}E\\
~~~~\neg{}D\lor{}F,\neg{}F\lor\neg{}G\lor{}C & \longrightarrow &
(\mathrm{learn~tab})\\

\neg{}A,B,\neg{}C,D,E,F,\neg{}G\parallel\neg{}A,A\lor{}B,A\lor\neg{}C,
\neg{}B\lor{}D,\neg{}D\lor{}E\\
~~~~\neg{}D\lor{}F,\neg{}F\lor\neg{}G\lor{}C,
G\lor\neg(H\equiv\lfloor\epsilon_x\in{}a\Rightarrow\epsilon_x\in{}a\rfloor) &
\longrightarrow & (\mathrm{unit~prop})\\

\neg{}A,B,\neg{}C,D,E,F,\neg{}G,\neg{}H\parallel\neg{}A,A\lor{}B,A\lor\neg{}C,
\neg{}B\lor{}D\\
~~~~\neg{}D\lor{}E,\neg{}D\lor{}F,\neg{}F\lor\neg{}G\lor{}C,
G\lor\neg{}H & \longrightarrow & (\mathrm{learn~tab})\\

\neg{}A,B,\neg{}C,D,E,F,\neg{}G,\neg{}H\parallel\neg{}A,A\lor{}B,A\lor\neg{}C,
\neg{}B\lor{}D\\
~~~~\neg{}D\lor{}E,\neg{}D\lor{}F,\neg{}F\lor\neg{}G\lor{}C,
G\lor\neg{}H,H\lor(I\equiv\lfloor\epsilon_x\in{}a\rfloor),\\
~~~~H\lor\neg{}I & \longrightarrow & (\mathrm{unit~prop})\\

\neg{}A,B,\neg{}C,D,E,F,\neg{}G,\neg{}H,I\parallel\neg{}A,A\lor{}B,A\lor\neg{}C,
\neg{}B\lor{}D\\
~~~~\neg{}D\lor{}E,\neg{}D\lor{}F,\neg{}F\lor\neg{}G\lor{}C,
G\lor\neg{}H,H\lor{}I,H\lor\neg{}I & \longrightarrow & (\mathrm{unsat})\\

\mathrm{unsat}\\\\

\mbox{with }\epsilon_x=\epsilon(x).\neg(x\in{}a\Rightarrow{}x\in{}a)
\end{array}$
\end{center}}
\caption{Example of Proof Using the Rewriting Theory}
\label{fig:rew}
\end{figure}
